Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-04-27 summary.Rmd
added subspace incoherence

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation"
authors: John Koo, Minh Tang, Michael Trosset
# author: John Koo
# author2: Minh Tang
# author3: Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output:
#   bookdown::pdf_document2:
#     citation_package: natbib
#     number_citations: yes
# output: html_document
# output: rticles::rss_article
fontsize: 12pt
# geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
bibliography: misc.bib
abstract: |
  We connect two random graph models, the Popularity Adjusted Block Model 
  (PABM) and the Generalized Random Dot Product Graph (GRDPG), demonstrating
  that a PABM is a GRDPG in which communities correspond to certain mutually
  orthogonal subspaces of latent positions. This insight leads to the
  construction of improved algorithms for community detection and parameter
  estimation with PABM. Using established asymptotic properties of Adjacency
  Spectral Embedding (ASE) for GRDPG, we derive asymptotic properties of these
  algorithms, including algorithms that rely on Sparse Subspace Clustering
  (SSC). We illustrate these properties via simulation.
---

@@

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{Sparse Subspace Clustering using LASSO \cite{jmlr-v28-wang13}.}
  \KwData{Adjacency matrix $A$, number of communities $K$, 
  hyperparameter $\lambda$}
  \KwResult{Community assignments $1, ..., K$}
    Find $V$, the matrix of eigenvectors of $A$ 
    corresponding to the $K (K + 1) / 2$ most positive 
    and the $K (K - 1) / 2$ most negative eigenvalues.\;
    \For {$i = 1, ..., n$} {
      Assign $v_i^\top$ as the $i^{th}$ row of $V$. 
      Assign $V_{-i} = \begin{bmatrix} 
      v_1 & \cdots & v_{i-1} & v_{i+1} & \cdots & v_n \end{bmatrix}^\top$.\;
      Solve the LASSO problem 
      $c_i = \arg\min_{\beta} 
      \frac{1}{2} ||v_i - V_{-i} \beta||_2^2 + \lambda ||\beta||_1$.\;
      Assign $\tilde{c}_i = \begin{bmatrix} 
      c_i^{(1)} & \cdots & c_i^{(i-1)} & 0 & c_i^{(i)} & \cdots & c_i^{(n-1)}
      \end{bmatrix}^\top$ such that the superscript is the index of 
      $\tilde{c}_i$.\;
    }
    Assign 
    $C = \begin{bmatrix} \tilde{c}_1 & \cdots & \tilde{c}_n \end{bmatrix}$.\;
    Compute the affinity matrix $B = \sqrt{n} (|C| + |C^\top|)$.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs (e.g., using edge 
    thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$. 
\end{algorithm}

@@

\begin{definition}[Inradius \cite{soltanolkotabi2012} \cite{jmlr-v28-wang13}]
The inradius of a convex body $\mathcal{P}$, denoted by $r(\mathcal{P})$, is 
defined as the radius of the largest Euclidean ball inscribed in $\mathcal{P}$.
In addition, $r(X)$ for data matrix $X$ with rows $x_i^\top$ represents 
the inradius of the symmetric convex hull of $X$. 
\end{definition}

\begin{lemma}
\label{lemma2}
Let $P_n = V_n \Lambda_n V_n^\top$ be the edge probability matrix of a PABM 
with $n$ vertices. Let $A_n \sim \Bernoulli(P_n)$ be the adjacency matrix 
drawn from $P_n$, and let $A_n = \hat{V}_n \hat{\Lambda}_n \hat{V}_n^\top$ be 
its approximate spectral decomposition using the $K (K+1) / 2$ most positive 
and $K (K-1) / 2$ most negative eigenvalues and their corresponding 
eigenvectors.Without loss of generality, order the rows of $V_n$ and 
$\hat{V}_n$ by community, and let $V_n^{(k)}$ and $\hat{V}_n^{(k)}$ be the 
$n_k \times K^2$ matrix of the rows of the $k^{th}$ community in $V_n$ and 
$\hat{V}_n$ respectively. 
Let $\mu_n^{(k)} = \mu(V_n^{(k)})$ be the subspace incoherence of $V_n^{(k)}$ 
as defined by \citeauthor{soltanolkotabi2012}. Then $\mu_n^{(k)} = 0$ 
with probability 1 $\forall n, k$.

Furthermore, Let $\hat{\mu}_n^{(k)} = \mu(\hat{V}_n^{(k)})$ using the 
extended definition of $\mu(\cdot)$ from \citeauthor{jmlr-v28-wang13}. Then

\begin{equation} \label{eq:mu-conv}
\hat{\mu}_n^{(k)} = 0
\end{equation}
\end{lemma}

*Proof*. 
For the first part, 
since $v_n^{(k)}$ is orthogonal to $v_n^{(l)}$ $\forall k \neq l$ 
(Lemma \ref{lemma1}), 
this falls directly out of Theorem 2.8 of \citeauthor{soltanolkotabi2012}.

For the second part, 
$\mu(\hat{V}_n^{(k)}) = 
\max_{v_n^{(i)} \in V_n^{(-k)}} ||U_n^{(k)} v_n^{(i)}||_\infty$ 
where the rows of $U_n^{(k)}$ consist of vectors lying on the subspace spanned 
by $V_n^{(k)}$. Since 
$u_n^{(i)} \in \text{span}(V_n^{(k)}) \implies u_n^{(i)} \perp v_n^{(j)}$ 
$\forall v_n^{(j)} \in V_n^{(-k)}$ by Theorem \ref{theorem2}, this quantity 
must be 0.

@@

\begin{remark}
Theorem 6 of \citeauthor{jmlr-v28-wang13} assume that each $||v_n^{(i)}|| = 1$, 
which scales each $r_n^{(k)} \leq 1$. This is not strictly necessary for 
the proof of Theorem \ref{theorem5} since each $\mu_n^{(k)} = 0$, so as long 
as the $k^{th}$ community spans its subspace, $r_n^{(k)} > 0 = \mu_n^{(k)}$.
\end{remark}

@@

*Case $k = l$*. $P^{(kk)}$ and $A^{(kk)}$ represent within-community edge 
probabilities and edges for community $k$.  
By definition, 
$P^{(kk)} = \lambda^{(kk)} (\lambda^{(kk)})^\top$. This implies that the 
signular value decomposition 
$P^{(kk)} = \sigma_{kk}^2 u^{(kk)} (u^{(kk)})^\top$ has one singular value and 
one pair of singular vectors ($P^{(kk)}$ is symmetric, so the left and right 
singular vectors are identical). Then $\lambda^{(kk)} = \sigma_{kk} u^{(kk)}$.  
Let $\hat{U}^{(kk)} \hat{\Sigma}^{(kk)} (\hat{U}^{(kk)})^\top$ be the singular 
value decomposition of $A^{(kk)}$, and let 
$\hat{\sigma}_{kk}^2 \hat{u}^{(kk)} (\hat{u}^{(kk)})^\top$ be its 
one-dimensional approximation. Define 
$\hat{\lambda}^{(kk)} = \hat{\sigma}_{kk} \hat{u}^{(kk)}$. Then 
$\hat{\lambda}^{(kk)}$ is the adjacency spectral embedding approximation of 
$\lambda^{(kk)}$.  
Then by Theorem 5 of \citeauthor{rubindelanchy2017statistical}, the adjacency 
spectral embedding $\hat{\lambda}^{(kk)}$ approximates $\lambda^{(kk)}$ at rate 
$\frac{(\log n_k)^c}{\sqrt{n_k}}$.

>>>>

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation"
authors: John Koo, Minh Tang, Michael Trosset
# author: John Koo
# author2: Minh Tang
# author3: Michael Trosset
output:
  pdf_document:
    keep_tex: true
    citation_package: natbib
    number_sections: yes
# output:
#   bookdown::pdf_document2:
#     citation_package: natbib
#     number_citations: yes
# output: html_document
# output: rticles::rss_article
fontsize: 12pt
# geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
bibliography: misc.bib
abstract: |
  We connect two random graph models, the Popularity Adjusted Block Model 
  (PABM) and the Generalized Random Dot Product Graph (GRDPG), demonstrating
  that a PABM is a GRDPG in which communities correspond to certain mutually
  orthogonal subspaces of latent positions. This insight leads to the
  construction of improved algorithms for community detection and parameter
  estimation with PABM. Using established asymptotic properties of Adjacency
  Spectral Embedding (ASE) for GRDPG, we derive asymptotic properties of these
  algorithms, including algorithms that rely on Sparse Subspace Clustering
  (SSC). We illustrate these properties via simulation.
---

@@

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{Sparse Subspace Clustering using LASSO \cite{jmlr-v28-wang13}.}
  \KwData{Adjacency matrix $A$, number of communities $K$, 
  hyperparameter $\lambda$}
  \KwResult{Community assignments $1, ..., K$}
    Find $V$, the matrix of eigenvectors of $A$ 
    corresponding to the $K (K + 1) / 2$ most positive 
    and the $K (K - 1) / 2$ most negative eigenvalues.\;
    Normalize $V \leftarrow \sqrt{n} V$.\;
    \For {$i = 1, ..., n$} {
      Assign $v_i^\top$ as the $i^{th}$ row of $V$. 
      Assign $V_{-i} = \begin{bmatrix} 
      v_1 & \cdots & v_{i-1} & v_{i+1} & \cdots & v_n \end{bmatrix}^\top$.\;
      Solve the LASSO problem 
      $c_i = \arg\min_{\beta} 
      \frac{1}{2} ||v_i - V_{-i} \beta||_2^2 + \lambda ||\beta||_1$.\;
      Assign $\tilde{c}_i = \begin{bmatrix} 
      c_i^{(1)} & \cdots & c_i^{(i-1)} & 0 & c_i^{(i)} & \cdots & c_i^{(n-1)}
      \end{bmatrix}^\top$ such that the superscript is the index of 
      $\tilde{c}_i$.\;
    }
    Assign 
    $C = \begin{bmatrix} \tilde{c}_1 & \cdots & \tilde{c}_n \end{bmatrix}$.\;
    Compute the affinity matrix $B = |C| + |C^\top|$.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs (e.g., using edge 
    thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$. 
\end{algorithm}

@@

\begin{definition}[Inradius \cite{soltanolkotabi2012, jmlr-v28-wang13}]
The inradius of a convex body $\mathcal{P}$, denoted by $r(\mathcal{P})$, is 
defined as the radius of the largest Euclidean ball inscribed in $\mathcal{P}$.
In addition, $r(X)$ for data matrix $X$ with rows $x_i^\top$ represents 
the inradius of the symmetric convex hull of $X$. 
\end{definition}

\begin{definition}[Subspace incoherence property \cite{jmlr-v28-wang13}]
\end{definition}

\begin{lemma}
\label{lemma2}
Let $\hat{V}$ be the eigenvectors of $A_n$ corresponding to the $K (K + 1) / 2$ most positive and $K (K - 1) / 2$ most negative eigenvalues such that the rows of $\hat{V}$ are ordered by community, and let $\hat{V}^{(k)}$ be the rows of the $k^{th}$ community in $\hat{V}$ and $\hat{V}^{(-k)}$ be the rows of $\hat{V}$ with the $k^{th}$ community omitted. Denote $(\hat{v}_i^{(k)})^\top$ as the rows of $\hat{V}$, $\hat{V}_{-i}^{(k)}$ as $\hat{V}^{(k)}$ with the $i^{th}$ row omitted, and $\mathcal{S}^{(k)}$ as the subspace spanned by $V^{(k)}$. Let $V$, $V^{(k)}$, $V^{(-k)}$, and $v_i^{(k)}$ be the corresponding values for $P_n$.

Let $\nu_{i}^{(k)} = \max\limits_\eta (\hat{v}_i^{(k)})^\top \eta - \frac{1}{2 \lambda} \eta^\top \eta$ subject to $||V_{-i}^{(k)} \eta||_\infty \leq 1$, and define the projected dual direction $w_{i}^{(k)}$ as $\mathbb{P}_{\mathcal{S}^{(k)}}(\nu_i^{(k)})$ normalized to length 1. Collect the projected dual directions into $W = \begin{bmatrix} w_1^{(k)} & \cdots & w_{n_k}^{(k)} \end{bmatrix}^\top$.

Define the subspace incoherence:

$$\mu_n^{(k)} = \mu(\hat{V}^{(k)}) = \max\limits_{v \in V^{(-k)}} ||W^{(k)} v||_\infty$$

Then $\forall k, n$,

\begin{equation} \label{eq:mu-conv}
\mu_n^{(k)} = 0
\end{equation}
\end{lemma}

*Proof*. 
Since by Theorem \ref{theorem2} each $\mathcal{S}^{(k)}$ are mutually orthogonal, any vector projected to 
$\mathcal{S}^{(k)}$ must be orthogonal to each row of $V^{(-k)}$. Therefore, $W^{(k)} v = 0$ $\forall v \in \mathcal{S}^{(-k)}$.

@@

\begin{remark}
Theorem 6 of \citeauthor{jmlr-v28-wang13} assume that each $||v_n^{(i)}|| = 1$, 
which scales each $r_n^{(k)} \leq 1$. This is not strictly necessary for 
the proof of Theorem \ref{theorem5} since each $\mu_n^{(k)} = 0$, so as long 
as the $k^{th}$ community spans its subspace, $a r_n^{(k)} > 0 = \mu_n^{(k)}$ 
$\forall a > 0$.
\end{remark}

@@

*Case $k = l$*. $P^{(kk)}$ and $A^{(kk)}$ represent within-community edge 
probabilities and edges for community $k$.  
By definition, 
$P^{(kk)} = \lambda^{(kk)} (\lambda^{(kk)})^\top$. This implies that the 
singular value decomposition 
$P^{(kk)} = \sigma_{kk}^2 u^{(kk)} (u^{(kk)})^\top$ has one singular value and 
one pair of singular vectors ($P^{(kk)}$ is symmetric, so the left and right 
singular vectors are identical). Then $\lambda^{(kk)} = \sigma_{kk} u^{(kk)}$.  
Let $\hat{U}^{(kk)} \hat{\Sigma}^{(kk)} (\hat{U}^{(kk)})^\top$ be the singular 
value decomposition of $A^{(kk)}$, and let 
$\hat{\sigma}_{kk}^2 \hat{u}^{(kk)} (\hat{u}^{(kk)})^\top$ be its 
one-dimensional approximation. Define 
$\hat{\lambda}^{(kk)} = \hat{\sigma}_{kk} \hat{u}^{(kk)}$. Then 
$\hat{\lambda}^{(kk)}$ is the adjacency spectral embedding approximation of 
$\lambda^{(kk)}$.  
Then by Theorem 5 of \citeauthor{rubindelanchy2017statistical}, the adjacency 
spectral embedding $\hat{\lambda}^{(kk)}$ approximates $\lambda^{(kk)}$ at rate 
$\frac{(\log n_k)^c}{\sqrt{n_k}}$.
